 

### 2021-06-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (@conditQuantifyingDeciduousnessTropical2000c, Meakem, *temperate REFS*). In this review, we focus on growing season conditions unless otherwise noted.

@@

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites. 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

@@

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, a that is optimal for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
Often, however, $T_{leaf}$ tracks $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020]. 
Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost to the environment (such as transpiration) [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Across species, inherent leaf $gs$ is generally greater in sun leaves than shade leaves, due to stomata density [Table 1]. However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--including increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, resulting in increased in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020]. Lastly, canopy structure and architecture adaptation can also influence canopy temperature  [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** 
Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. Comparing upper canopies, temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however considering a broader range of forests, this might not be always the case (Fig. `r fig_NEON_vertical`). At a specie-specific level, leaf boundary resistance is lower for conifers than broad-leaf species, resulting in higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018]. Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].

@@

Exposed canopy environment induces traits in sun leaves that are better adaptated to heat and water stress than shade leaves through, for instance, being smaller, hairier, with greater vein density and LMA [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], as well as strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*]. These traits [see Table 1] collectively dissapate heat and regulate $T_{leaf}$-photosynthetic balance. Whereas, shade leaves have traits that enable rapid light capture in the diffused-light environment with sunflecks, for example greater SLA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013a]. In open canopies where light is comparitevely homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

Biochemical protection against foliage light and heat damage also increases with irradiance.
Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. This is also important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a]. N and P are essential for photosynthetic and metabolic processes along with $VAZ$, as they contribute to enzyme formation and function, such as in Rubisco. Particularly, $N_{area}$ increases in the direction of new leaf growth, where it also optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018]. Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed canopies, making these species more advantageous [*REF*].

@@

Deciduousness as a trait in tropical forests is an adaptation to conserve water under stress. Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter. Among tropical species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000c]. In deciduous-temperate forests, spatio-temporal leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al. 2013; @seiwaChangesLeafPhenology1999].Therefore, deciduous leaf-loss or -gain can alter biophysical and $T_{leaf}$ gradient [@rey-sanchez_spatial_2016].

@@

Exposed sun-leaves, though equipped with thermoregulatory traits, are more constrained by $gs$ than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sunleaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006]. The diurnal $gs$ pattern and resulting $A_{max}$ varies between species; for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].

@@

Further, leaf thermal tolerances are more closely adapted to microenvironment than macroclimate [@feeley_thermal_2020; Slot et al. 2020]. Particularly, in an Australian arid region, *Acacia papyrocarpa* (Benth.) trees showed higher composite climate stress and greater $T_{50}$ in lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019]. However, more research is required to fully understand the relationship between leaf thermal tolerance and microenvironments.

@@

A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy. Other studies have observed no difference in $Q_{10}$ across species vertical profiles and in seedling sun vs. shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008]. Although, @bolstad_foliar_1999 found that temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased, [@bolstad_foliar_1999](*see their Fig. 1*). (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*) and also found evidence of acclimation, but there remained a declining trend of respiration with topographical elevation.
Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] also showed no variation within canopy positions in Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].

@@

**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality. There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)*

Biophysical conditions can alter tree-water dynamics and growth patterns. Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013]. Whereas, greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. Altered precipitation regimes causing drought are more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020]. However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

Tolerance to multiple biophysical and biological stressors can differ among tree size classes, having been shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to multiple stressors [@niinemetsResponsesForestTrees2010a]. Autotrophic respiration also increases with stand age that declines NPP, whereas GPP increases with associated morphological and anatomical complexity with age [@IncreasingRespiratoryCosts]. These difference can alter tree growth patterns in response to biophysical conditions. Particularly, higher temperatures are shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014]. Evidence in tropical forests also shows that higher nighttime temperatures are associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supress growth [*Clark 2010; 2013; Feeley et al. 2007*]. .

@@

The complex interactions of the biophysical environment and biological--leaf traits, metabolic processes and distribution of trees--are highly interconneced, resulting in feedback loops to another. Therefore, climate-driven changes have implications on forests at many levels.

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts[Bennett et al. 2015]. Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Increase in such canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].

Canopy structure affects understory species composition and soil fertility which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]. Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, despite buffering dampening extreme effects, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]. This imposes an increasing threat to fauna in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].

>>>>

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In this review, we focus on growing season conditions unless otherwise noted.

@@

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites. 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

@@

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, a optimal for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
Often, however, $T_{leaf}$ tracks $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020]. 
Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost to the environment (such as transpiration) [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Across species, inherent leaf $gs$ is generally greater in sun leaves than shade leaves, due to stomata density [Table 1]. However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--including increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, resulting in increased in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ and thermal capacitance can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020]. Lastly, canopy structure and architecture adaptation can also influence canopy temperature  [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** 
Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. Upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig. `r fig_NEON_vertical`). At a specie-specific level, leaf boundary resistance is lower for conifers than broad-leaf species, resulting in higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018]. Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].

@@

Exposed canopy environment induces traits in sun leaves that are better adaptated to heat and water stress than shade leaves where, for instance, being smaller, hairier, with greater vein density and LMA [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], as well as strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*]. These traits [see Table 1] collectively aid heat dissapation and regulate $T_{leaf}$-photosynthetic balance. Whereas, shade leaves have traits that enable rapid light capture in the diffused-light environment with sunflecks, for example greater SLA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013a]. In open canopies where light is comparitevely homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

Biochemical protection against foliage light and heat damage also increases with irradiance.
Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. This is also important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a]. Elemental traits, N and P though differ with soil conditions, $N_{area}$ typically increases in the direction of new leaf growth, where it also optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018]. Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed canopies, making these species more advantageous [*REF*].

@@

Deciduousness as a trait in tropical forests is an adaptation to conserve water under stress. Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter. Among tropical species individuals that can be deciduous, the fraction of deciduous trees tends to increase across tree size classes, with understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000c; Meakem]. In deciduous-temperate forests, spatio-temporal leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al. 2013; @seiwaChangesLeafPhenology1999].Therefore, deciduous leaf-loss or -gain can alter biophysical and $T_{leaf}$ gradient [@rey-sanchez_spatial_2016].

@@

Exposed sun-leaves, though equipped with thermoregulatory traits, are more constrained by $gs$ than shade leaves that are buffered within-canopy [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006]. Even so, $T_{opt}$ of $gs$ appears to be similar in sun and shade leaves [Slot et al. 2019]. This influences leaf thermal sensitivity [Perez and Feeley, 2018], but exact mechanisms remain unclear. Inter-specifically, diurnal $gs$ pattern and resulting $A_{max}$ varies where, for instance, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].

@@

Leaf thermal tolerances are more closely adapted to microenvironment than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Particularly, in an Australian arid region, *Acacia papyrocarpa* (Benth.) trees showed higher composite climate stress and greater $T_{50}$ in lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019]. However, more research is required to fully understand the relationship between leaf thermal tolerance and microenvironments.

@@

A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy. Other studies have observed no difference in $Q_{10}$ across species vertical profiles and in seedling sun vs. shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008]. Although, in @bolstad_foliar_1999, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased, [@bolstad_foliar_1999](*see their Fig. 1*). (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*) with evidence of acclimation and a declining trend of respiration with topographical elevation.
Similary, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] did not differ within canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].

@@

**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality. There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors. Tolerance of more than one biological and biophyical stressors can increase with tree height and light, partly due to non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a]. However, growth also declines as trees age. This is partly because autotrophic respiration increases with age that declines NPP, whereas due to increase in morphological and anatomical complexity, GPP also increases [@IncreasingRespiratoryCosts]. Vertical biophysical conditions are bound to have different influenceS on size class and growth rates due to these existing demographic differences.

Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013]. Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. Altered precipitation regimes cause droughts, that are more  detrimental to larger trees than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020]. However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

Biophysical conditions can alter growth rates, but can do so differently between size classes through influencing existing demographic differences. Particulary evidence in in mesic temperate and subtropical eucalyptus forests shows that higher temperatures primarily reduced tree growth through altering water availability  [@bowmanWarmerWorldWill2014]. Whereas in tropical forests higher nighttime temperatures was associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees [*Clark 2010; 2013; Feeley et al. 2007*], perhaps due to increase in nighttime metabolic respiration that could supress growth.

@@

The complex interwoven relations between the biophysical environment and biological factors--leaf traits, metabolic processes and distribution of trees-- are likely to produce amplifications and feedback loops in a warming world, with implications for forests on many levels.

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of forest microenvironments [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly as larger trees suffer more during droughts[Bennett et al. 2015]. Further, increasing global temperatures show that canopies of temperate and tropical forests are already functioning close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Such increase in canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than in temperate forests [Perez and Feeley, 2018].

Canopy structure affects understory species composition and soil fertility, which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]. Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy. Even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]. This imposes an increasing threat to flora and fauna in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
